---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit only README.Rmd! -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r, echo = FALSE}
library(Ryacas)
```


# Ryacas #

[![Build Status](https://travis-ci.org/r-cas/ryacas.svg?branch=master)](https://travis-ci.org/r-cas/ryacas)
[![Build status](https://ci.appveyor.com/api/projects/status/c8fsb1dvj5gmh703/branch/master?svg=true)](https://ci.appveyor.com/project/r-cas/ryacas/branch/master)
[![JOSS status](https://joss.theoj.org/papers/69b3947f8900504a25aa36f65d14500b/status.svg)](https://joss.theoj.org/papers/69b3947f8900504a25aa36f65d14500b)
[![DOI](https://zenodo.org/badge/36067045.svg)](https://zenodo.org/badge/latestdoi/36067045)

Ryacas is an [R](https://www.r-project.org/) interface to
the free [yacas](http://www.yacas.org) Computer Algebra
System.  Ryacas allows one to send unprocessed yacas strings and 
certain other R objects to
yacas process from R and get back the result. It also has
facilities for manipulating yacas strings and R expressions
destined for yacas processing.

It can be used for arbitrary-precision arithmetic, symbolic math, ASCII
pretty printing and translating R to TeX. 

## Install from GitHub ##

To build and install from Github using R 3.3.0 (or later) and the R `devtools` package 1.11.0 (or later) run this command from within `R`:

```
devtools::install_github("r-cas/ryacas", 
                         build_opts = c("--no-resave-data", "--no-manual"))
```

You can also install the package without vignettes if needed as follows:

```
devtools::install_github("r-cas/ryacas")
```


## Online info ##

For vignettes, overview, pointers to additional information, installation
instructions and a sample session see <http://r-cas.github.io/ryacas/>. 

Yacas documentation can be found at http://yacas.readthedocs.org/.

## Contribute, issues, and support ##

Please use the issue tracker at <https://github.com/r-cas/ryacas/issues> 
if you want to notify us of an issue or need support.
If you want to contribute, please either create an issue or make a pull request.

## Brief examples ##

Below we show a few examples. 
We highly recommend reading the "[Getting started](http://r-cas.github.io/ryacas/articles/getting-started.html)" vignette (and the other vignettes) for a more thorough introduction to the package.

There are two interfaces: a high-level interface that 
makes `yacas` objects work similar to `R` objects, 
and a low-level interface where the user can write `yacas` code 
and get results as strings or as `R` expressions. Below, we demonstrate both.

### High-level interface

A brief example with a polynomial is:
```{r}
x <- ysym("x")
p <- x^2+x-6
p
y_fn(p, "Factor")
p %>% y_fn("Factor")
p %>% as_r()
```

A small matrix example follows:

```{r}
A <- outer(0:3, 1:4, "-") + diag(2:5)
a <- 1:4
B <- ysym(A)
B
solve(B)
B[2, 3] <- "x"
B
b <- ysym(a)
b[1] <- "x"
b
B %*% b
t(B)
B[, 2:3]
```


### Low-level interface

Returning strings with `yac_str()`:

```{r}
yac_str("x+x+x+x")
yac_str("Factor(x^2+x-6)")
yac_str("D(x) x^2+x-6")
```

Returning `R` expressions with `yac_expr()`:

```{r}
yac_expr("x+x+x+x")
eval(yac_expr("x+x+x+x"), list(x = 4))
yac_expr("Factor(x^2+x-6)")
yac_expr("D(x) x^2+x-6")
```

Using functions easier (using [`magrittr`](https://cran.r-project.org/package=magrittr)'s pipe, `%>%`):

```{r}
"x^2+x-6" %>% y_fn("Factor") %>% yac_str()
"x^2+x-6" %>% y_fn("D(x)") %>% yac_expr()
```

Solving equations (removes the `x==` from `yacas` with the `y_rmvars()` function):

```{r}
sol <- yac_str("Solve(x^2+x-6 == 0, x)")
sol
sol %>% y_rmvars() %>% yac_str()
sol %>% y_rmvars() %>% yac_expr()
sol %>% y_rmvars() %>% yac_expr() %>% eval()
```

And output in TeX:

```{r}
"3/4 + Pi/8" %>% y_fn("Simplify") %>% y_fn("TeXForm") %>% yac_str()
```

And arbitrary precision (see also the "[Arbitrary-precision arithmetic
](http://r-cas.github.io/ryacas/articles/arbitrary-precision.html)" vignette):

```{r}
yac_str("N(Pi, 50)")
```



## Yacas ##

The package contains the yacas distribution. The development of the yacas source code is available at <https://github.com/grzegorzmazur/yacas/> . For more information on yacas see <http://www.yacas.org/> or the documention directly at <https://yacas.readthedocs.io/>.

---

Mikkel Meyer Andersen, mikl at math dot aau dot dk  
Rob Goedman, goedman at mac dot com  
Gabor Grothendieck, ggrothendieck at gmail dot com  
Søren Højsgaard, sorenh at math dot aau dot dk  
Ayal Pinkus, apinkus at xs4all dot nl  
Grzegorz Mazur, teoretyk at gmail dot com  

