---
title: "Using `Ryacas` in automatic generation of exams in R"
author: "Mikkel Meyer Andersen and Søren Højsgaard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Ryacas in automatic generation of exams in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(Ryacas)
set.seed(1) # For reproducibility
```

The [`exams` package](https://cran.r-project.org/package=exams) offer a wide range of ways of generating exams. See e.g. the many [examples](http://www.r-exams.org/templates/).

`Ryacas` can be used in connection with the [`exams` package](https://cran.r-project.org/package=exams) in multiple ways. The obvious ones are for symbolic mathematics and for turning `R` code into TeX.

In this vignette we demonstrate both by two simple examples. First, we make exercises where the student must find roots in a quadratic polynomial. We generate exercises where the roots are two different integers and the polynomial are written in simplified form using `Ryacas` (in principle this can be solved purely numeric as well, but the solution is less elegant). 
Then we make exercises in simple factorisations, e.g. "Bring $9 x ^{2} + 30 x y + 25 y ^{2}$ on the form
$(a \cdot x + b \cdot y)^2$ for positive $a$ and $b$.".
Both examples show how symbolic math and TeX from the `Ryacas` package are used.

```{r, include=FALSE}
library(exams)

out_dir <- tempdir()
```

# Roots in polynomials

This is the exams `Rmd` file:

```{r, comment=NA, echo=FALSE}
exam_file <- system.file("exams/poly-roots.Rmd", package = "Ryacas")
fl <- readLines(exam_file)
cat(paste0(fl, "\n"))
```

This can then be compiled using e.g. `exams::exams2html`, `exams::exams2pandoc` or any of the other `exams::exams2*` functions to obtain something like:

```{r, include=FALSE}
out_nm <- "out-exams-poly"

exams::exams2pandoc(exam_file, n = 2, type = "markdown", dir = out_dir, name = out_nm)
```


Giving first sample:

```{r, comment=NA, echo=FALSE}
fl <- readLines(file.path(out_dir, paste0(out_nm, "1.markdown")))
cat(paste0(fl, "\n"))
```


And second sample:

```{r, comment=NA, echo=FALSE}
fl <- readLines(file.path(out_dir, paste0(out_nm, "2.markdown")))
cat(paste0(fl, "\n"))
```


# Factorisation

This is the exams `Rmd` file:

```{r, comment=NA, echo=FALSE}
exam_file <- system.file("exams/poly-factorisation.Rmd", package = "Ryacas")
fl <- readLines(exam_file)
cat(paste0(fl, "\n"))
```

This can then be compiled using e.g. `exams::exams2html`, `exams::exams2pandoc` or any of the other `exams::exams2*` functions to obtain something like:

```{r, include=FALSE}
out_nm <- "out-exams-factorisation"

exams::exams2pandoc(exam_file, n = 2, type = "markdown", dir = out_dir, name = out_nm)
```


Giving first sample:

```{r, comment=NA, echo=FALSE}
fl <- readLines(file.path(out_dir, paste0(out_nm, "1.markdown")))
cat(paste0(fl, "\n"))
```


And second sample:

```{r, comment=NA, echo=FALSE}
fl <- readLines(file.path(out_dir, paste0(out_nm, "2.markdown")))
cat(paste0(fl, "\n"))
```

