---
title: "Getting started"
author: "Mikkel Meyer Andersen and Søren Højsgaard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(Ryacas)
```


# Introduction

`Ryacas` makes the `yacas` computer algebra system available from
within `R`. The name `yacas` is short for "Yet Another Computer Algebra
System". The `yacas` program 
is developed by Ayal Pinkhuis and others, 
and is available at <http://www.yacas.org/> for various
platforms. There is a comprehensive documentation (300+ pages) of
`yacas` available at <https://yacas.readthedocs.io> and the
documentation contains many examples.

## Note

This version of `Ryacas` is somewhat different to previous versions of `Ryacas` 
because we have tried to make the interface a lot simpler.

The old version of `Ryacas` is available as a legacy version called `Ryacas0` at <https://github.com/mikldk/ryacas0/> with documentation directly available at <https://mikldk.github.io/ryacas0/>.

# Interfaces to `yacas`

The naming principle governing `Ryacas` functions is as follows:

* `yac_*(x)` functions evaluate/run `yacas` command `x`; the result varies depending on which of the functions used (see below)
* `y_*(x)` various utility functions (not involving calls to `yacas`)

There are two interfaces to `yacas`: a low-level (see the "The low-level interface" vignette) and a high-level (see the "The high-level (symbol) interface" vignette). 
The low-level is highly customisable, but also requires more work with text strings. 
The high-level is easier dealing with vectors and matrices, but may also be less computationally efficient and less flexible.

Below, we will demonstrate both interfaces and refer to the other vignettes 
for more information.

A short summary of often-used `yacas` commands are found 
at the end of this vignette.
A short summary of often-used low-level `Ryacas` functions are found 
at the end of the "The low-level interface" vignette, 
and a short summary of often-used high-level `Ryacas` functions are found 
at the end of the "The high-level (symbol) interface" vignette.



## The low-level interface

The low-level interface consists of these two main functions:

* `yac_str(x)`: Evaluate `yacas` command `x` (a string) and get result as **string/character**.
* `yac_expr(x)`: Evaluate `yacas` command `x` (a string) and get result as an **`R` expression**.

Note, that the `yacas` command `x` is a string and must often be built op using `paste()`/`paste0()`. 
Examples of this will be shown in multiple examples below. 


### Examples

```{r}
eq <- "x^2 + 4 + 2*x + 2*x"
yac_str(eq) # No task was given to yacas, so we simply get the same returned
yac_str(paste0("Simplify(", eq, ")"))
yac_str(paste0("Factor(", eq, ")"))
yac_expr(paste0("Factor(", eq, ")"))
yac_str(paste0("TeXForm(Factor(", eq, "))"))
```

Instead of the pattern `paste0("Simplify(", eq, ")")` etc., there exists a 
helper function `y_fn()` that does this:

```{r}
y_fn(eq, "Factor")
yac_str(y_fn(eq, "Factor"))
yac_str(y_fn(y_fn(eq, "Factor"), "TeXForm"))
```

As you see, there are a lot of nested function calls. That can be avoided by using `magrittr`'s pipe `%>%` (automatically available with `Ryacas`) together with the helper function `y_fn()`:

```{r}
eq %>% y_fn("Factor")
eq %>% y_fn("Factor") %>% yac_str()
eq %>% y_fn("Factor") %>% y_fn("TeXForm") %>% yac_str()
```

The polynomial can be evaluated for a value of $x$ by calling `yac_expr()` instead of `yac_str()`:
```{r}
yac_str(paste0("Factor(", eq, ")"))
expr <- yac_expr(paste0("Factor(", eq, ")"))
expr
eval(expr, list(x = 2))
```



## The high-level interface

The high-level interface consists of the main function `yac_symbol()` and 
often the helper function `as_r()` will be used to get back an `R` object (expression, matrix, vector, ...).


### Examples

Before we had `eq` as a text string. We now make a `yac_symbol()` from that:
```{r}
eqy <- yac_symbol(eq)
eqy
as_r(eqy)
eqy %>% y_fn("Factor") # Notice how we do not need to call yac_str()/yac_expr()
```
Notice how the printing is different from before.

We start with a small matrix example:

```{r}
A <- outer(0:3, 1:4, "-") + diag(2:5)
a <- 1:4
B <- yac_symbol(A)
B
b <- yac_symbol(a)
b
```

Notice how they are printed using `yacas`'s syntax.

We can apply `yacas` functions using `y_fn()`:

```{r}
y_fn(B, "Transpose")
y_fn(B, "Inverse")
y_fn(B, "Trace")
```

Some standard `R` commands are available (see the section "`Ryacas` high-level reference" at the end of the "The high-level (symbol) interface" vignette):

```{r}
A %*% a
B %*% b
t(A)
t(B)
A[, 2:3]
B[, 2:3]
A %*% solve(A)
B %*% solve(B)
```

Next we will demonstrate matrix functionality using the [Hilbert matrix](https://en.wikipedia.org/wiki/Hilbert_matrix) 
\[
H_{{ij}}={\frac{1}{i+j-1}} 
\]
In `R`'s `solve()` help file there is code for generating it:
```{r, eval = FALSE}
hilbert <- function(n) { 
  i <- 1:n
  H <- 1 / outer(i - 1, i, "+")
  return(H)
}
```

To avoid floating-point issues (see the "Arbitrary-precision arithmetic" vignette), 
we instead generate just the denominators as a stanard `R` matrix:
```{r}
hilbert_den <- function(n) { 
  i <- 1:n
  H <- outer(i - 1, i, "+")
  return(H)
}
Hden <- hilbert_den(4)
Hden
H <- 1/Hden
H
```

To use `Ryacas`'s high-level interface, we use the function `yac_symbol()` 
that converts the matrix to `yacas` representation and automatically calls 
`yac_str()` when needed.
Furthermore, it enables standard `R` functions such as subsetting with `[`, `diag()`, `dim()` and others.

```{r}
Hyden <- yac_symbol(Hden)
Hyden
Hy <- 1/Hyden
Hy
```

Notice how the printing is different from `R`'s printing.

We can then to a number of things with the `yac_symbol()`.

```{r}
as_r(Hy) # now floating-point and the related problems
diag(Hy)
Hy[upper.tri(Hy)]
Hy[1:2, ]
dim(Hy)
A <- Hy
A[lower.tri(A)] <- "x"
A
as_r(A)
eval(as_r(A), list(x = 999))
```






# `yacas` reference

Below are some `yacas` functions. A more elaborate reference is available at <https://yacas.readthedocs.io/>:

* General
    + `Expand(x)`: Expand an expression
    + `Factor(x)`: Factorise an expression
    + `Simplify(x)`: Simplify an expression
    + `Solve(expr, var)` solve an equation (refer to the `Ryacas` function `y_rmvars()`)
    + `Variables()`: List `yacas` variables
* Calculus:
    + `D(x) expr`: Take the derivative of `expr` with respect to `x`
    + `HessianMatrix(function, var)`: Create the Hessian matrix
    + `JacobianMatrix(function, var)`: Create the Jacobian matrix
    + `Limit(n, a) f(n)`: Limit of `f(n)` for `n` going towards `a` (e.g. `Infinity` or `0`)
    + `Sum(k, a, b, f(k))`: Sum of `f(k)` for `k` from `a` to `b`.
* Output
    + `TeXForm(x)`: Get a $\LaTeX$ representation of an expression
    + `PrettyForm(x)`: Print a prettier ASCII representation of an expression
* Linear algebra
    + `Inverse(A)`: Inverse of a matrix
    + `Transpose(A)`: Transpose of a matrix
    + `A * B`: Matrix multiplication (and not as `R`'s `%*%`)


