---
title: "Getting started"
author: "Mikkel Meyer Andersen and Søren Højsgaard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(Ryacas)
```


# Introduction

`Ryacas` makes the `yacas` computer algebra system available from
within `R`. The name `yacas` is short for "Yet Another Computer Algebra
System". The `yacas` program 
is developed by Ayal Pinkhuis (who is also the maintainer)
and others, and is available at <http://www.yacas.org/> for various
platforms. There is a comprehensive documentation (300+ pages) of
`yacas` (also available at
<http://www.yacas.org/>) and the
documentation contains many examples.

# Example 1: using character vectors

```{r}
eq <- "x^2 + 4 + 2*x + 2*x"
yacas(eq)
yacas(paste0("Simplify(", eq, ")"))
yacas(paste0("Factor(", eq, ")"))
```

Or by using the corresponding `R` functions directly:
```{r}
Simplify(yacas(eq))
Factor(yacas(eq))
```

Or course the result can be saved before calling the helper `R` functions:

```{r}
yeq <- yacas(eq)
Simplify(yeq)
Factor(yeq)
```

And getting results out:

```{r}
res <- Factor(yeq)
as.character(res)
TeXForm(res)
```

To include inline, e.g. $`r TeXForm(yeq)` = `r TeXForm(res)`$.


# Example 2: using `Sym()` 

```{r}
xs <- Sym("x")
eqs <- xs^2 + 4 + 2*xs + 2*xs
Simplify(eqs)
Factor(eqs)
res <- Factor(eqs)
as.character(res)
TeXForm(res)
```


