---
title: "The high-level (symbol) interface"
author: "Mikkel Meyer Andersen and Søren Højsgaard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The high-level (symbol) interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(Ryacas)
```

A short summary of often-used `yacas` commands are found in the section "`yacas` reference" in the "Getting started" vignette.
A short summary of `Ryacas`'s high-level functions are found in the section "`Ryacas` high-level reference" at the end of this document.

# Introduction

Start with a base symbol what can either be:

* A `yacas` command, e.g. `x`, `2*a` or something similar
* An `R` matrix or vector.

Here, we keep it simple. 

```{r}
x <- ysym("x")
2*x^2 - 5
c(-2, 5)*x
c(2*x, -x^3)
as_r(c(-2, 5)*x) # or yac_expr(c(-2, 5)*x)
```

Then consider an `R` matrix and vector:

```{r}
A <- outer(0:3, 1:4, "-") + diag(2:5)
a <- 1:4
A
a
```

They are now considered `yacas`-enabled:

```{r}
B <- ysym(A)
B
as_r(B)
b <- ysym(a)
b
as_r(b)
```

Notice how they are printed using `yacas`'s syntax.

We can apply `yacas` functions using `y_fn()`:

```{r}
y_fn(B, "Transpose")
y_fn(B, "Inverse")
y_fn(B, "Trace")
```

Standard `R` commands are available (see the section "`Ryacas` high-level reference" at the end of this document):

```{r}
A %*% a
B %*% b
t(A)
t(B)
exp(B)
as_r(exp(B))
A[, 2:3]
B[, 2:3]
A[upper.tri(A)] <- 1
B[upper.tri(B)] <- 1
A
B
2*A - A
2*B - B
A %*% solve(A)
B %*% solve(B)
solve(A %*% t(A))
solve(B %*% t(B))
solve(A, a)
solve(B, b)
```

We can also assign a `yacas` variable, but remember that this may be difficult to distinguish:

```{r}
yac_str("W") # Get variable W if exists, or else just a symbol
yac_str("Variables()") # ...or list variables
B
yac_assign(B, "W") # assign B in R to W in yacas
yac_str("W") # Get variable W if exists, or else just a symbol
yac_str("Variables()") # ...or list variables
yac_silent("Clear(W)")
yac_str("Variables()") # List variables
yac_str("W") # Get variable W if exists, or else just a symbol
```


# Simplify and output to TeX

There are additional functions available:

* `simplify()`
* `tex()`

To demonstrate these and some additional benefit, we exploit `yacas`'s symbolic availabilities.

```{r}
D <- diag(4) %>% ysym()
D
D <- D/2
D
D[2:3, 1] <- "d"
D[3, 4] <- "2*d + 2"
D
D %>% solve()
D %>% solve() %>% simplify()
D %>% solve() %>% simplify() %>% tex()
```

\[
`r D %>% solve() %>% simplify() %>% tex()`
\]

`yacas` has a `Simplify()` function. This is made available via a `simplify()` function that also includes a time-out that prevents `yacas` in making the `R` session hang, but it requires that the `unix` package is available.
The default `timeout` value used when `unix` is available is `2` seconds.

# Derivatives

We illustrate using the example in <https://mikl.dk/post/2019-pizza-frozen-yogurt-goebner/>:

```{r}
L <- ysym("x^2 * (y/4) - a*(3*x + 3*y/2 - 45)")
L
```

We can consider one variable only:

```{r}
deriv(L, "x")
Hessian(L, "x")
```

Or multiple variables:

```{r}
deriv(L, c("x", "y", "a"))
H <- Hessian(L, c("x", "y", "a"))
H
as_r(H)
eval(as_r(H), list(x = 2, y = 2, a = 2))
```

The Jacobian is taken on a vector function denoted by many functions:

```{r}
L2 <- ysym(c("x^2 * (y/4) - a*(3*x + 3*y/2 - 45)", 
                   "x^3 + 4*a^2")) # just some function
L2
Jacobian(L2, "x")
Jacobian(L2, c("x", "y", "a"))
```


# Solving equations

Say we want to find roots of a polynomial. We use the generic `solve(a, b, ...)` function. 

Note the conventions are as follows:

* Linear system of equations (works as `R`'s `solve()` as demonstrated above):
  + When `a` is a matrix and `b` not provided, this finds the inverse of `a`.
  + When `a` is a matrix and a vector `b` is provided, the linear system of equations is solved.
* Else:
  + `solve(a, b)`: find roots of `a` for variable `b`, i.e. yacas `Solve(a == 0, b)`
  + `solve(a, b, v)`: find solutions to `a == b` for variable `v`, i.e. yacas `Solve(a == b, v)`

```{r}
xs <- ysym("x")
poly <- xs^2 - xs - 6
poly
zeroes <- solve(poly, "x") # Solve(x^2 - x - 6 == 0, x)
zeroes
tex(zeroes)
zeroes %>% y_rmvars()
```

We can also find values of `x` where the polynomial equals another constant. 
If we were working with strings via the low-level interface it would be easy via `paste()`, but as we are working with `ysym()`'s we use the `solve()` function directly:

```{r}
solve(poly, 3, "x") # Solve(x^2 - x - 6 == 3, x)
solve(poly, 3, "x") %>% tex()
```

\[
`r solve(poly, 3, "x") %>% tex()`
\]

## Solving a system of equations



```{r}
x <- ysym("x")
y <- ysym("y")
lhs <- c(3*x*y - y, x)
rhs <- c(-5*x, y+4)
```

$$\begin{align}
`r tex(lhs[[1]])` &= `r tex(rhs[[1]])` \\
`r tex(lhs[[2]])` &= `r tex(rhs[[2]])`
\end{align}$$


```{r}
sol <- solve(lhs, rhs, c("x", "y"))
sol
sol_vals <- lapply(seq_len(nrow(sol)), function(sol_no) {
  y_rmvars(sol[sol_no, ])
})
sol_vals
sol_envir <- lapply(sol_vals, function(l) {
  list(x = as_r(l[1]), y = as_r(l[2]))
})
sol_envir
do.call(rbind, lapply(seq_along(sol_envir), function(sol_no) {
  sol_val <- sol_envir[[sol_no]]
  data.frame(sol_no = sol_no,
             eq_no = seq_along(sol_val),
             lhs = eval(as_r(lhs), sol_val),
             rhs = eval(as_r(rhs), sol_val))
}))
```


# `Ryacas` high-level reference

Principle:

* `ysym(x)` converts `x` to a `yac_symbol` that automatically runs `yacas` when needed. `x` can both be a text string with `yacas` commands or an `R` vector/matrix.
* `as_r(x)`: Is used to convert the `yac_symbol` back to an `R` representation.
* `y_fn(x, fn, ...)`: Apply a `yacas` function `fn` to the `yac_symbol` `x`, i.e. `fn(x, ...)`; note that this is evaluated immediately when `x` is a `yac_symbol` as opposed to when `x` is a string

Reference:

The following functions work with `yac_symbol`s.

* `yac_*()` functions (see the "Getting started" vignette)
  + `yac_str()`: Return `yacas` string
  + `yac_expr()`: Return `R` expression
  + `yac_silent()`: Do something silently
  + `yac_assign()`: Assign a variable
* Other
  + `simplify(x, timeout = 2)`: Try `yacas`'s `Simplify()` function. When the `unix` package is available, the `timeout` (in seconds), stops trying after that amount of time to avoid making the `R` process hang.
  + `tex()`: Convert 
  + `y_fn(x, fn, ...)`: Apply a `yacas` function `fn` to the `yac_symbol` `x`, i.e. `fn(x, ...)`
  + `y_rmvars(x)`: Remove variable names in `x`
  + Derivatives:
    - `deriv(expr, vars)`: takes derivative of `yac_symbol` `expr` with respect to `vars`
    - `Jacobian(expr, vars)`: finds Jacobian of `yac_symbol` `expr` (usually a vector of expressions) with respect to `vars`
    - `Hessian(expr, vars)`: finds Hessian matrix of `yac_symbol` `expr` with respect to `vars`
  + Other
    - `integrate()`
    - `sum()`
    - `lim()`
* Standard `R` that has been implemented for `yac_symbol`s:
  + `print()`
  + `c()`
  + `dim()`
  + `[` getter
  + `[<-` setter
  + `[[` getter
  + `%*%` matrix/vector multiplication
  + `diag()` getter
  + `diag<-()` setter
  + `upper.tri()` getter
  + `lower.tri()` getter
  + `t()`
  + `solve()` (see above and in help page)
  + Ops: `+`, `-`, `*`, `/`, `^`
  + Math functions: `sin()`, `cos()`, `tan()`, 
                    `asin()`, `acos()`, `atan()`, 
                    `asinh()`, `acosh()`, `atanh()`, 
                    `exp()`, `log()`, `sqrt()`
