---
title: "The high-level (symbol) interface"
author: "Mikkel Meyer Andersen and Søren Højsgaard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The high-level (symbol) interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(Ryacas)
```

A short summary of often-used `yacas` commands are found in the section "`yacas` reference" in the "Getting started" vignette.
A short summary of `Ryacas`'s high-level functions are found in the section "`Ryacas` high-level reference" at the end of this document.

# Introduction

Start with a base symbol what can either be:

* A `yacas` command, e.g. `x`, `2*a` or something similar
* An `R` matrix or vector.

Here, we keep it simple. Consider an `R` matrix and vector:

```{r}
A <- outer(0:3, 1:4, "-") + diag(2:5)
a <- 1:4
A
a
```

They are now considered `yacas`-enabled:

```{r}
B <- yac_symbol(A)
B
b <- yac_symbol(a)
b
```

Notice how they are printed using `yacas`'s syntax.

We can apply `yacas` functions using `y_fn()`:

```{r}
y_fn(B, "Transpose")
y_fn(B, "Inverse")
y_fn(B, "Trace")
```

Standard `R` commands are available (see the section "`Ryacas` high-level reference" at the end of this document):

```{r}
A %*% a
B %*% b
t(A)
t(B)
exp(B)
as_r(exp(B))
A[, 2:3]
B[, 2:3]
A[upper.tri(A)] <- 1
B[upper.tri(B)] <- 1
A
B
2*A - A
2*B - B
A %*% solve(A)
B %*% solve(B)
solve(A %*% t(A))
solve(B %*% t(B))
```

We can also assign a `yacas` variable, but remember that this may be difficult to distinguish:

```{r}
yac_str("W") # Get variable W if exists, or else just a symbol
yac_str("Variables()") # ...or list variables
B
yac_assign(B, "W") # assign B in R to W in yacas
yac_str("W") # Get variable W if exists, or else just a symbol
yac_str("Variables()") # ...or list variables
yac_silent("Clear(W)")
yac_str("Variables()") # List variables
yac_str("W") # Get variable W if exists, or else just a symbol
```


There are additional functions available:

* `simplify()`
* `tex()`

To demonstrate these and some additional benefit, we exploit `yacas`'s symbolic availabilities.

```{r}
D <- diag(4) %>% yac_symbol()
D
D <- D/2
D
D[2:3, 1] <- "d"
D[3,4] <- "2*d + 2"
D
D %>% solve()
D %>% solve() %>% simplify()
D %>% solve() %>% simplify() %>% tex()
```

\[
`r D %>% solve() %>% simplify() %>% tex()`
\]


`yacas` has a `Simplify()` function. This is made available via a `simplify()` function that also includes a time-out that prevents `yacas` in making the `R` session hang, but it requires that the `unix` package is available.
The default `timeout` value used when `unix` is available is `2` seconds.


# `Ryacas` high-level reference

Principle:

* `yac_symbol(x)` converts `x` to a `yac_symbol` that automatically runs `yacas` when needed. `x` can both be a text string with `yacas` commands or an `R` vector/matrix.
* `as_r(x)`: Is used to convert the `yac_symbol` back to an `R` representation.
* `y_fn(x, fn, ...)`: Apply a `yacas` function `fn` to the `yac_symbol` `x`, i.e. `fn(x, ...)`

Reference:

The following functions work with `yac_symbol`s.

* `yac_*()` functions (see the "Getting started" vignette)
  + `yac_str()`: Return `yacas` string
  + `yac_expr()`: Return `R` expression
  + `yac_silent()`: Do something silently
  + `yac_assign()`: Assign a variable
* Other
  + `simplify(x, timeout = 2)`: Try `yacas`'s `Simplify()` function. When the `unix` package is available, the `timeout` (in seconds), stops trying after that amount of time to avoid making the `R` process hang.
  + `tex()`: Convert 
* Stanard `R` that has been implemented for `yac_symbol`s:
  + `print()`
  + `dim()`
  + `[` getter
  + `[<-` setter
  + `%*%` matrix/vector multiplication
  + `diag()` getter
  + `diag<-()` setter
  + `upper.tri()` getter
  + `lower.tri()` getter
  + `t()`
  + `solve()` (only matrix inverse)
  + Ops: `+`, `-`, `*`, `/`, `^`
  + Math functions: `sin()`, `cos()`, `tan()`, 
                    `asin()`, `acos()`, `atan()`, 
                    `asinh()`, `acosh()`, `atanh()`, 
                    `exp()`, `log()`, `sqrt()`
