---
title: "The low-level interface"
author: "Mikkel Meyer Andersen and Søren Højsgaard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The low-level interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(Ryacas)
```

The low-level interface consists of these two main functions:

* `yac_str(x)`: Evaluate `yacas` command `x` (a string) and get result as **string/character**.
* `yac_expr(x)`: Evaluate `yacas` command `x` (a string) and get result as an **`R` expression**.

Note, that the `yacas` command `x` is a string and must often be built op using `paste()`/`paste0()`. 
Examples of this will be shown in multiple examples below. 

A short summary of often-used `yacas` commands are found in the section "`yacas` reference" in the "Getting started" vignette.
A short summary of `Ryacas`'s low-level functions are also found in the section "`Ryacas` low-level reference" at the end of this document.

# Example 1: Simple algebra

First, consider this polynomial:

```{r}
eq <- "x^2 + 4 + 2*x + 2*x"
```

Now, perform `yacas` operations, and get result as string/character:

```{r}
yac_str(eq) # No task was given to yacas, so we simply get the same returned
yac_str(paste0("Simplify(", eq, ")"))
yac_str(paste0("Factor(", eq, ")"))
yac_str(paste0("TeXForm(Factor(", eq, "))"))
```

\[
`r yac_str(paste0("TeXForm(Factor(", eq, "))"))`
\]

Instead of the pattern `paste0("Simplify(", eq, ")")` etc., there exists a 
helper function `y_fn()` that does this:

```{r}
y_fn(eq, "Simplify")
yac_str(y_fn(eq, "Simplify"))
yac_str(y_fn(eq, "Factor"))
yac_str(y_fn(y_fn(eq, "Factor"), "TeXForm"))
```

As you see, there are a lot of nested function calls. That can be avoided by using `magrittr`'s pipe `%>%` (automatically available with `Ryacas`) together with the helper function `y_fn()`:

```{r}
eq %>% y_fn("Simplify")
eq %>% y_fn("Simplify") %>% yac_str()
eq %>% y_fn("Factor") %>% yac_str()
eq %>% y_fn("Factor") %>% y_fn("TeXForm") %>% yac_str()
```

Below, we will stick to the standard way of calling the functions, and not using the pipe, but now it has been demonstrated if the user prefers that way.

We will not use the pipe operator below, but just demonstrate its usage.

Now, again perform `yacas` operations, but get result as an `R` expression, e.g. for continued computations:

```{r}
eq
eq %>% yac_expr() # Alternative to "yac_expr(eq)"
cmd <- eq %>% y_fn("Factor")
cmd
e <- yac_expr(cmd)
e
eval(e, list(x = 2))
```

# Example 2: Linear algebra

To work with matrices and vectors, you need to realise that `yacas` and `R` has different 
ways of representing these objects. `yacas` represents vectors as a list, and a matrix as a list of lists (each list is a row).

## Simple example

You can work with these directly. Here illutrated with vectors:

```{r}
cmd <- "2*{x, x^2, x^3}"
cmd %>% yac_str()
e <- cmd %>% yac_expr()
e
eval(e, list(x = 1.5))
```

And then illutrated with matrices. First with purely numeric contents:

```{r}
cmd <- "{{1, 2}, {3, 4}}"
yac_str(cmd)
y_print(cmd) # Convenience function for prettier print
e <- cmd %>% yac_expr()
e
eval(e)
```

Also in $\LaTeX$:

```{r}
cmd %>% y_fn("TeXForm") %>% yac_str()
```


But it also works with symbolic contents:

```{r}
cmd1 <- paste0("a * ", cmd, "")
cmd2 <- cmd1 %>% y_fn("Inverse")
cmd2
cmd2 %>% yac_str()
cmd2 %>% y_fn("TeXForm") %>% yac_str()
```

\[
`r cmd2 %>% y_fn("TeXForm") %>% yac_str()`
\]

```{r}
paste0(cmd2, "*", cmd1) %>% 
  y_fn("Simplify") %>% 
  yac_str()
e2 <- cmd2 %>% yac_expr()
eval(e2, list(a = 2.2))
```

## Using `R`'s character matrices

The above is fine when writing `yacas` vectors and matrices by hand. 
But often one would want to exploit `R`'s convenient functions to work with matrices. 

The central idea to make this possible is to work with `R` character matrices. 
We provide two helper functions to go back and forth between `R` and `yacas`:

* `as_y(x)`: Convert `R` character matrix `x` to a `yacas` representation
* `as_r(x)`: Convert a `yacas` representation `x` to a `R` character matrix

Below, we illustrate the usage of both functions.

First, we create a character matrix using `R`:

```{r}
Achr <- matrix(0, nrow = 3, ncol = 3)
diag(Achr) <- 1
Achr[2, 3] <- "a"
Achr[1, 3] <- "a"
Achr
```

Note how this is a character matrix. If we want to find it's inverse symbolically using `yacas`, it must first be represented as a `yacas` matrix:

```{r}
Ayac <- Achr %>% as_y()
Ayac
```

Now, we can find the inverse:

```{r}
cmd <- Ayac %>% y_fn("Inverse")
cmd %>% yac_str()
```

A nicer representation can be obtained in (at least) four ways:

```{r}
way1 <- cmd %>% yac_str()
way1 %>% y_print()
way2 <- cmd %>% y_fn("TeXForm") %>% yac_str()
way2
way3 <- cmd %>% y_fn("PrettyForm") %>% yac_str()
way3 %>% cat() # Result of PrettyForm() must be printed
way4 <- cmd %>% yac_str()
way4
way4 %>% as_r()
way4 %>% as_r() %>% print(quote = FALSE)
```

Say we want to subset it to only consider a submatrix.
To do that, we can use `R`'s facilities:

```{r}
A_inv_yac <- way4 %>% as_r()
Bchr <- A_inv_yac[2:3, 2:3]
Bchr
Bchr %>% as_y()
Bchr %>% as_y() %>% 
    y_fn("Inverse") %>% 
    yac_str() %>% 
    as_r()
```

# `yacas` variables

`yacas` also has variables. They are assigned by `:=`. 

Consider this example:

```{r}
yac_str("poly := (x-3)*(x+2)")
```

If the output is not necessary, it can be suppressed by using `yac_silent()` instead of `yac_str()`:

```{r}
yac_silent("poly := (x-3)*(x+2)")
```

We can now list `yacas` variables (`I` is the imaginary unit):

```{r}
yac_str("Variables()")
```

```{r}
yac_str("Expand(poly)")
"poly" %>% y_fn("Expand") %>% yac_str()
```


# Sums

```{r}
yac_str("Sum(k, 0, n, a^k)")
``` 


# Limits

```{r}
cmd <- "Limit(n, Infinity) (1+(1/n))^n"
yac_str(cmd)
yac_expr(cmd)
yac_str("Limit(h, 0) (Sin(x+h)-Sin(x))/h")
``` 


# Solving equations

Say we want to find roots of `poly`. First note that the equality in `yacas` is `==`. Then_

```{r}
cmd <- "Solve(poly == 0, x)"
cmd %>% yac_str()
```

If we want the solution without the variable name `x` and the equality symbol `==`, we can use `Ryacas` helper function `y_rmvars()`:

```{r}
cmd
cmd %>% y_rmvars() %>% yac_str()
cmd %>% y_rmvars() %>% yac_expr()
```

We can also use `y_fn()`:

```{r}
"poly == 0" %>% y_fn("Solve", "x")
"poly == 0" %>% y_fn("Solve", "x") %>% y_rmvars() %>% yac_str()
```


# Case: gradient and Hessian for a function

Say we have a function:

```{r}
f <- function(x, y) 2*x^2 + 3*y + y*x^2
f_body <- body(f)
f_body
f_body_chr <- as.character(as.expression(body(f)))
f_body_chr
# or:
# f_body_chr <- "2 * x^2 + 3 * y + y * x^2"
# f <- function(x, y) NULL
# body(f) <- parse(text = f_body_chr, keep.source = FALSE)
```

The gradient can be found using `yacas` (`D(x) expr` is the derivative of `expr` with respect to `x`):

```{r}
cmd_g <- paste0("{ D(x) ", f_body_chr, ", D(y) ", f_body_chr, " }")
cmd_g
g_body <- yac_expr(cmd_g)
g_body
g <- function(x, y) NULL
body(g) <- g_body
g
g(2, 4)
```

The Hessian matrix can also be found using `yacas`:

```{r}
cmd_H <- paste0("HessianMatrix(", f_body_chr, ", {x, y})")
cmd_H
H_body <- yac_expr(cmd_H)
H_body
H <- function(x, y) NULL
body(H) <- H_body
H
H(2, 4)
```

# `Ryacas` low-level reference

Principle:

* `yac_*(x)` functions evaluate/run `yacas` command `x`; the result varies depending on which of the functions used
* `y_*(x)` various utility functions (not involving calls to `yacas`)

Reference:

* Evaluate `yacas` expressions
    + `yac(x, rettype = c("str", "expr", "silent"))`: Evaluate `yacas` command `x` (a string) and get result determined by `rettype` (default `"str"`).
    + `yac_expr(x)`: Evaluate `yacas` command `x` (a string) and get result as an **`R` expression**.
    + `yac_silent(x)`: Evaluate `yacas` command `x` (a string) silently; useful for creating `yacas` variables.
**string/character**.
    + `yac_str(x)`: Same as `yac_expr()`, but get result as **string/character**.
* Helper functions
    + `as_y(x)`: Convert `R` character matrix `x` to a `yacas` representation
    + `as_r(x)`: Convert a `yacas` representation `x` to an `R` object
    + `y_fn(x, fn, ...)`: Helper function to prepare a call for `yacas`, e.g. `y_fn("x^2 - 1", "Factor")` is gives `"Factor(x^2 - 1)"`; `...` are additional arguments: `y_fn(x, fn, ...)` gives `fn(x, ...)`
    + `y_rmvars(x)`: Removes variables such that `{x == 2}` instead gets `{2}`; remember to call `yacas` with e.g. `yac_str()` or `yac_expr()`
    + `y_print(x)`: Pretty print yacas strings, e.g. a yacas matrix
* Lower level functions
    + `yac_core(x)`: Evaluate `yacas` command `x` (a string) and get both result and side effects; used in the implementation of `yac_expr()`, `yac_silent()`, and `yac_str()`
