---
title: "The structure of the concentration and covariance matrix in a simple state-space model"
author: "Mikkel Meyer Andersen and Søren Højsgaard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The structure of the concentration and covariance matrix in a simple state-space model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=F}
library(Ryacas)
library(tidyverse)
library(Matrix)
```

Set output width:

```{r}
get_output_width()
set_output_width(120)
get_output_width()
```



## Autoregression ($AR(1)$)

Consider $AR(1)$ process: $x_i = a x_{i-1} + e_i$ where $i=1,2,3$ and where $x_0=e_0$. Isolating error terms gives that
$$
 e = L_1 x
$$
where $e=(e_0, \dots, e_3)$ and $x=(x_0, \dots x_3)$ and where $L_1$ has the form
```{r}
syacas("L1 := {{ 1,  0,  0, 0}, 
               {-a,  1,  0, 0}, 
               { 0, -a,  1, 0}, 
               { 0,  0, -a, 1}}")
```

If error terms have variance $1$ then $\mathbf{Var}(e)=L \mathbf{Var}(x) L'$ so the covariance matrix $V1=\mathbf{Var}(x) = L^- (L^-)'$ while the concentration matrix is $K=L L'$

```{r}
syacas("K1 := Simplify(L1 * Transpose(L1))")
syacas("V1 := Simplify(Inverse(K1))")
```

```{r}
K1 = yacas("K1 := Simplify(L1 * Transpose(L1))")
V1 = yacas("V1 := Simplify(Inverse(K1))")
```

```{r, results="asis"}
K1 <- Sym("K1") ## Uskønt...
V1 <- Sym("V1")
cat(
  "\\begin{align} K_1 &= ", TeXForm(K1), " \\\\ 
                  V_1 &= ", TeXForm(V1), " \\end{align}", sep = "")
```

## Dynamic linear model

Augument the $AR(1)$ process above with $y_i=b x_i + u_i$. Then
$(e,u)$ can be expressed in terms of $(x,y)$ as
$$
(e,u) = L_2(x,y)
$$
where
```{r}
syacas("L2 := { { 1,  0,  0,  0, 0, 0, 0}, 
                {-a,  1,  0,  0, 0, 0, 0},   
                { 0, -a,  1,  0, 0, 0, 0}, 
                { 0,  0, -a,  1, 0, 0, 0},
                { 0, -b,  0,  0, 1, 0, 0}, 
                { 0,  0, -b,  0, 0, 1, 0}, 
                { 0,  0,  0, -b, 0, 0, 1}}")
```

```{r}
syacas("K2 := Simplify(L2 * Transpose(L2))")
syacas("V2 := Simplify(Inverse(K2))")

K2s <- Sym("K2")
V2s <- Sym("V2")
```

```{r, results="asis"}
cat(
  "\\begin{align} K_2 &= ", TeXForm(K2s), " \\\\ 
                  V_2 &= ", TeXForm(V2s), " \\end{align}", sep = "")
```


## Numerical evalation in R

```{r}
sparsify <- function(x){Matrix::Matrix(x, sparse=T)}

alpha <- .5
beta <- -.3

## AR(1)
N <- 3
L1 <- diag(1, 1 + N)
L1[cbind(1+(1:N), 1:N)] <- -alpha
K1 <- L1 %*% t(L1)
V1 <- solve(K1)
sparsify(K1)
sparsify(V1)

## Dynamic linear models
N <- 3
L2 <- diag(1, 1 + 2*N)
L2[cbind(1+(1:N), 1:N)] <- -alpha
L2[cbind(1 + N + (1:N), 1 + 1:N)] <- -beta
L2
K2 <- L2 %*% t(L2)
V2 <- solve(K2)
sparsify(K2)
sparsify(V2)
```

Comparing with results calculated by yacas:

```{r}
yres_V2 <- Eval(V2s, list(a = alpha, b = beta))
yres_V2_mat <- matrix(unlist(yres_V2), nrow = 7, ncol = 7)

V2
yres_V2_mat
V2 - yres_V2_mat
all.equal(V2, yres_V2_mat)
```

