---
title: "User-defined yacas rules"
author: "Mikkel Meyer Andersen and Søren Højsgaard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-defined yacas rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(Ryacas)
```


# Included rules

`yacas` comes with a number of rules all defined in the `yacas` directory of the installed package:

```{r}
system.file(package = "Ryacas", "yacas")
```

For example in the `sums.rep` folder, a number of rules for sums are defined in the `code.ys` file.

As an example, the fact that
\[
  \sum_{k = 1}^n (2k-1) = n^2
\]
is defined in `yacas` as
```
SumFunc(_k,1,_n,2*_k-1, n^2 );
```
and the geometric sum is defined as
```
SumFunc(_k,0,_n,(r_IsFreeOf(k))^(_k), (1-r^(n+1))/(1-r) );
```

These can be verified:
```{r}
yac_str("Sum(i, 1, m, 2*i-1)")
yac_str("Sum(i, 0, m, 2^i)")
```

There are also rules in `yacas` that are able to let the user change some limits of some sums, e.g. for the geometric sum:

```{r}
yac_str("Sum(i, 1, m, 2^i)")
```

# Custom rules

But what about changing the limit of the first sum? I.e. instead of 
\[
  \sum_{k = 1}^n (2k-1) = n^2
\]
then know that
\[
  \sum_{k = 0}^n (2k-1) = -1 + \sum_{k = 1}^n (2k-1) = n^2 - 1 .
\]
But what does `yacas` say?
```{r}
yac_str("Sum(i, 0, m, 2*i-1)")
```

We can then add our own rule by:
```{r}
yac_silent("SumFunc(_k,0,_n,2*_k-1, n^2 - 1)")
```

And then try again:

```{r}
yac_str("Sum(i, 0, m, 2*i-1)")
```

A good source of inspiration for writing custom rules is reading the included rules, 
but there is a lot to programming in `yacas` and we refer to `yacas`'s documentation, 
specifically the chapter [Programming in Yacas](https://yacas.readthedocs.io/en/latest/programming_in_yacas/).

